---
title: "Week 6: Advanced R Graphics and ggplot2"
output:
  revealjs::revealjs_presentation:
    theme: white
    center: true
    transition: none
    incremental: true
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(knitr)
library(readr)
library(dplyr)
knitr::opts_chunk$set(echo = TRUE)
```

# Advanced R Graphics

## NCAA Basketball data
We will use data from the NCAA basketball tournament from 2011 - 2016.
```{r, message = F}
hoops <- read_csv('http://www.math.montana.edu/ahoegh/teaching/stat408/datasets/TourneyDetailedResults.csv')
hoops_2011 <- hoops %>% filter(Season >= 2011)
hoops_2011
```

## Compute annual averages
```{r}
points <- hoops_2011 %>%  group_by(Season) %>% 
  summarise(Win.Points = mean(Wscore), Lose.Points = mean(Lscore), 
            Win.3Pt = mean(Wfgm3), Lose.3pt = mean(Lfgm3)) 
head(points)
```

## Plot types: Points 
```{r, echo = F}
plot(points$Lose.Points~points$Season,ylim=c(0,max(points$Lose.Points)), ylab='Average Points Scored', xlab='Season', main='Average Points Scored by Losing Teams \n in NCAA Tournament',cex.lab=.8, cex.main = .8)
```

## Plot types: Points Code
```{r, eval = F}
plot(points$Lose.Points~points$Season,
     ylim=c(0,max(points$Lose.Points)), 
     ylab='Average Points Scored', xlab='Season', 
     main='Average Points Scored by Losing Teams \n in NCAA Tournament',
     cex.lab=.8, cex.main = .8)
```

## Plot types: Bars
```{r, echo=F}
plot(points$Lose.Points~points$Season,ylim=c(0,max(points$Lose.Points)),
     ylab='Average Points Scored', xlab='Season', type='h',lwd=3,
     main='Average Points Scored by Losing Teams \n in NCAA Tournament',cex.lab=.8, cex.main = .8)
```

## Plot types: Bars Code
```{r, eval=F}
plot(points$Lose.Points~points$Season,
     ylim=c(0,max(points$Lose.Points)),
     ylab='Average Points Scored', 
     xlab='Season', type='h',lwd=3,
     main='Average Points Scored by Losing Teams \n in NCAA Tournament',
     cex.lab=.8, cex.main = .8)
```

## Plot Types: Lines
```{r, echo=F}
plot(points$Lose.Points~points$Season,
     ylim=c(0,max(points$Lose.Points)),
     ylab='Average Points Scored', xlab='Season', type='n',
     main='Average Points Scored by Losing Teams \n in NCAA Tournament',cex.lab=.8, cex.main = .8)
lines(points$Lose.Points~points$Season,col='red',lwd=1)
lines(points$Win.Points~points$Season,col='blue',lwd=1,lty=2)
```

## Plot Types: Lines Code
```{r, mysize=TRUE, size='\\tiny', eval=F}
plot(points$Lose.Points~points$Season,
     ylim=c(0,max(points$Lose.Points)),
     ylab='Average Points Scored', xlab='Season', 
     type='n',
     main='Average Points Scored by Losing Teams \n in NCAA Tournament',
     cex.lab=.8, cex.main = .8)
lines(points$Lose.Points~points$Season,col='red',lwd=1)
lines(points$Win.Points~points$Season,col='blue',lwd=1,lty=2)
```

## Lines and Legends
```{r, echo=F}
plot(points$Lose.Points~points$Season,ylim=c(0,max(points$Win.Points)),
     ylab='Average Points Scored', xlab='Season', type='n',
     main='Average Points Scored in NCAA Tournament',cex.lab=.8, cex.main = .8)
lines(points$Lose.Points~points$Season,col='red',lwd=1)
lines(points$Win.Points~points$Season,col='blue',lwd=1,lty=2)
legend('bottomright',legend=c('Winning Team','Losing Team'),col=c('blue','red'),
       lwd=c(1,1), lty=c(2,1), cex = .7)
```

## Lines and Legends: Code
```{r, eval=F}
plot(points$Lose.Points~points$Season,
     ylim=c(0,max(points$Win.Points)),
     ylab='Average Points Scored', xlab='Season', 
     type='n',
     main='Average Points Scored in NCAA Tournament',
     cex.lab=.8, cex.main = .8)
lines(points$Lose.Points~points$Season,col='red',lwd=1)
lines(points$Win.Points~points$Season,col='blue',lwd=1,lty=2)
legend('bottomright',legend=c('Winning Team','Losing Team'),
       col=c('blue','red'),
       lwd=c(1,1), lty=c(2,1), cex = .7)
```

## Points
```{r, echo=F}
plot(points$Lose.Points~points$Season,ylim=c(0,max(hoops_2011$Wscore)),
     ylab='Average Points Scored', xlab='Season', type='l', lwd=1,col='red',
     main='Average Points Scored in NCAA Tournament',cex.lab=.8, cex.main = .8)
lines(points$Win.Points~points$Season,col='blue',lwd=1,lty=2)
points(hoops_2011$Wscore~hoops_2011$Season,pch=16,col=rgb(0,0,.5,.1))
legend('bottomright',legend=c('Average Winning Team','Average Losing Team', 'Individual Winning Team'),col=c('blue','red',rgb(0,0,.5)),
       lwd=c(1,1,NA), lty=c(2,1,NA), pch=c(NA,NA,16), cex=.7)
```

## Points: Code
```{r, eval=F}
plot(points$Lose.Points~points$Season,
     ylim=c(0,max(hoops_2011$Wscore)),
     ylab='Average Points Scored', xlab='Season', 
     type='l', lwd=1,col='red',
     main='Average Points Scored in NCAA Tournament',
     cex.lab=.8, cex.main = .8)
lines(points$Win.Points~points$Season,col='blue',lwd=1,lty=2)
points(hoops_2011$Wscore~hoops_2011$Season,pch=16,col=rgb(0,0,.5,.1))
legend('bottomright',
       legend=c('Average Winning Team','Average Losing Team', 'Individual Winning Team'),
       col=c('blue','red',rgb(0,0,.5)),
       lwd=c(1,1,NA), lty=c(2,1,NA), pch=c(NA,NA,16), cex=.7)
```


## Annotation
```{r, echo=F}
plot(points$Lose.Points~points$Season,
          ylim=c(0,max(hoops_2011$Wscore)),
     ylab='Average Points Scored', xlab='Season', type='l', lwd=1,col='red',
     main='Average Points Scored in NCAA Tournament',cex.lab=.8, cex.main = .8)
lines(points$Win.Points~points$Season,col='blue',lwd=1,lty=2)
points(hoops_2011$Wscore~hoops_2011$Season,pch=16,col=rgb(0,0,.5,.1))
text(2014.5,40,'Shot Clock Shortened', cex=.6)
arrows(x0=2015, y0=42, x1=2016, y1=70,  length=0.1, lwd=2)
legend('bottomright',legend=c('Average Winning Team','Average Losing Team', 'Individual Winning Team'),col=c('blue','red',rgb(0,0,.5)),
       lwd=c(1,1,NA), lty=c(2,1,NA), pch=c(NA,NA,16), cex=.6)
```

## Annotation: Code
```{r, mysize=TRUE, size='\\tiny', eval=F}
plot(points$Lose.Points~points$Season,
          ylim=c(0,max(hoops_2011$Wscore)),
     ylab='Average Points Scored', xlab='Season', type='l', lwd=1,col='red',
     main='Average Points Scored in NCAA Tournament',cex.lab=.8, cex.main = .8)
lines(points$Win.Points~points$Season,col='blue',lwd=1,lty=2)
points(hoops_2011$Wscore~hoops_2011$Season,pch=16,col=rgb(0,0,.5,.1))
text(2014.5,40,'Shot Clock Shortened', cex=.6)
arrows(x0=2015, y0=42, x1=2016, y1=70,  length=0.1, lwd=2)
legend('bottomright',
       legend=c('Average Winning Team','Average Losing Team', 'Individual Winning Team'),
       col=c('blue','red',rgb(0,0,.5)),
       lwd=c(1,1,NA), lty=c(2,1,NA), pch=c(NA,NA,16), cex=.6)
```

## Axes
```{r, echo=F}
plot(points$Lose.Points~points$Season,
     ylim=c(0,max(hoops_2011$Wscore)), axes=F,
     ylab='Average Points Scored', xlab='Season', type='b',pch=16, lwd=1,col='red',
     main='Average Points Scored in NCAA Tournament', cex.main=.75, cex.lab=.75)
lines(points$Win.Points~points$Season,col='blue',lwd=1,lty=2,type='b',pch=17)
legend('bottomright',legend=c('Average Winning Team','Average Losing Team'),col=c('blue','red'),
       lwd=1, lty=c(2,1),pch=c(17,16), cex=.7)
```

## Axes: Code
```{r, mysize=TRUE, size='\\tiny', eval=F}
plot(points$Lose.Points~points$Season,
     ylim=c(0,max(hoops_2011$Wscore)), axes=F,
     ylab='Average Points Scored', xlab='Season',
     type='b',pch=16, lwd=1,col='red',
     main='Average Points Scored in NCAA Tournament',
     cex.main=.75, cex.lab=.75)
lines(points$Win.Points~points$Season,col='blue',
      lwd=1,lty=2,type='b',pch=17)
legend('bottomright',
       legend=c('Average Winning Team','Average Losing Team'),
       col=c('blue','red'),
       lwd=1, lty=c(2,1),pch=c(17,16), cex=.7)
```

## Axes
```{r, echo=F}
plot(points$Lose.Points~points$Season,
     ylim=c(0,max(hoops_2011$Wscore)), axes=F,
     ylab='Average Points Scored', xlab='Season',
     type='b',pch=16, lwd=1,col='red',
     main='Average Points Scored in NCAA Tournament',
     cex.main=.75, cex.lab=.75)
lines(points$Win.Points~points$Season,
      col='blue',lwd=3,lty=2,type='b',pch=17)
legend('bottomleft',
       legend=c('Average Winning Team','Average Losing Team'),
       col=c('blue','red'),
       lwd=1, lty=c(2,1),pch=c(17,16), cex=.7)
axis(4)
axis(1, at = 2011:2016,
     labels=c('10-11','11-12','12-13','13-14','14-15','15-16'))
```

## Axes
```{r, eval=F}
plot(points$Lose.Points~points$Season,
     ylim=c(0,max(hoops_2011$Wscore)), axes=F,
     ylab='Average Points Scored', xlab='Season',
     type='b',pch=16, lwd=1,col='red',
     main='Average Points Scored in NCAA Tournament',
     cex.main=.75, cex.lab=.75)
lines(points$Win.Points~points$Season,
      col='blue',lwd=3,lty=2,type='b',pch=17)
legend('bottomleft',
       legend=c('Average Winning Team','Average Losing Team'),
       col=c('blue','red'),
       lwd=1, lty=c(2,1),pch=c(17,16), cex=.7)
axis(4)
axis(1, at = 2011:2016,
     labels=c('10-11','11-12','12-13','13-14','14-15','15-16'))
```

## Axes Box
```{r, echo=F}
plot(points$Lose.Points~points$Season,
     ylim=c(0,max(hoops_2011$Wscore)), axes=F,
     ylab='Average Points Scored', xlab='Season',
     type='b',pch=16, lwd=1,col='red',
     main='Average Points Scored in NCAA Tournament',
     cex.main=.75, cex.lab=.75)
lines(points$Win.Points~points$Season,
      col='blue',lwd=3,lty=2,type='b',pch=17)
legend('bottomleft',
       legend=c('Average Winning Team','Average Losing Team'),
       col=c('blue','red'),
       lwd=1, lty=c(2,1),pch=c(17,16), cex=.7)
axis(4)
axis(1, at = 2011:2016,
     labels=c('10-11','11-12','12-13','13-14','14-15','15-16'))
box()
```

## Axes Box: Code
```{r, eval=F}
plot(points$Lose.Points~points$Season,
     ylim=c(0,max(hoops_2011$Wscore)), axes=F,
     ylab='Average Points Scored', xlab='Season',
     type='b',pch=16, lwd=1,col='red',
     main='Average Points Scored in NCAA Tournament',
     cex.main=.75, cex.lab=.75)
lines(points$Win.Points~points$Season,
      col='blue',lwd=3,lty=2,type='b',pch=17)
legend('bottomleft',
       legend=c('Average Winning Team','Average Losing Team'),
       col=c('blue','red'),
       lwd=1, lty=c(2,1),pch=c(17,16), cex=.7)
axis(4)
axis(1, at = 2011:2016,
     labels=c('10-11','11-12','12-13','13-14','14-15','15-16'))
box()
```

## Exercise: Advanced Plotting
Use the Seattle Housing Data Set [http://math.montana.edu/ahoegh/teaching/stat408/datasets/SeattleHousing.csv](http://math.montana.edu/ahoegh/teaching/stat408/datasets/SeattleHousing.csv) to create an interesting graphic, include informative titles, labels, and add an annotation.

```{r}
seattle_in <- read_csv('http://math.montana.edu/ahoegh/teaching/stat408/datasets/SeattleHousing.csv')
```

## Solution: Advanced Plotting
```{r, echo=F,fig.align='center'}
hist(seattle_in$price,prob=T,breaks="FD", ylab='', col='forestgreen',xlab='Sales Price (million $)', main='Houses Sold in Seattle',
     axes=F)
axis(1, at = c(0,500000,1000000,2500000,4000000,5500000,7000000), labels =c('0','.5','1','2.5','4','5.5','7'))

arrows(x0=2500000, y0=1.5e-6, x1=1500000, y1=.5e-6,  length=0.1, lwd=2)
text(2500000,1.8e-6,'Most homes sell for \n less than one million',cex=.8)
```

## Solution: Advanced Plotting Code
```{r, eval=F,mysize=TRUE, size='\\tiny'}
hist(seattle_in$price,prob=T,breaks="FD", ylab='',
     col='forestgreen',
     xlab='Sales Price (million $)', 
     main='Houses Sold in Seattle', axes=F)
axis(1, at = c(0,500000,1000000,2500000,4000000,5500000,
               7000000), 
     labels =c('0','.5','1','2.5','4','5.5','7'))

arrows(x0=2500000, y0=1.5e-6, x1=1500000, y1=.5e-6, 
       length=0.1, lwd=2)
text(2500000,1.8e-6,
     'Most homes sell for \n less than one million',cex=.8)
```

## Superimposed Plots
```{r, echo=F}
plot(density(hoops_2011$Wscore),xlab='points',ylab='', main='histogram and superimposed density curve \n for points scored by winning team',lwd=3, cex.main=.8, cex.lab=.8)
hist(hoops_2011$Wscore,add=T,probability = T)
```

## Superimposed Plots: Code
```{r, eval = F}
plot(density(hoops_2011$Wscore),xlab='points',ylab='',
     main='histogram and superimposed density curve \n for points scored by winning team',
     lwd=3, cex.main=.8, cex.lab=.8)
hist(hoops_2011$Wscore,add=T,probability = T)
```

## Expression / Text

```{r, echo=T}
plot(density(hoops_2011$Wscore),
     ylab=expression(beta[2]),xlab='',
     main='Examples with Expression',axes=F, type='n')
box()
text(70,.023, expression(sum(theta[i]^2, i=1, n)),cex=2)
```



# ggplot2

## ggplot2 Overview
Why ggplot2?

### Advantages of ggplot2

- consistent underlying grammar of graphics (Wilkinson, 2005)
- plot specification at a high level of abstraction
- very flexible
- theme system for polishing plot appearance

## Grammar of Graphics
The basic idea: independently specify plot building blocks and combine them to create just about any kind of graphical display you want.

Building blocks of a graph include:

- data
- aesthetic mapping
- geometric object
- statistical transformations
- faceting

## ggplot2 VS Base Graphics

Compared to base graphics, ggplot2

- is more verbose for simple / canned graphics
- is less verbose for complex / custom graphics
- does not have methods (data should always be in a data.frame)
- uses a different system for adding plot elements

## Aesthetic Mapping

Aesthetics are things that you can see. Examples include:

>- position (i.e., on the x and y axes)
>- color ("outside" color)
>- fill ("inside" color)
>- shape (of points)
>- linetype
>- size

Aesthetic mappings are set with the aes() function.

## Geometric Objects (geom)

Geometric objects are the actual marks we put on a plot. Examples include:

>- points (`geom_point`)
>- lines (`geom_line`)
>- boxplot (`geom_boxplot`)

A plot must have at least one geom; there is no upper limit. You can add a geom to a plot using the `+` operator

## Graphical Primitives/ ggplot
```{r, fig.align='center', fig.height=2.75}
graph.a <- ggplot(data = hoops_2011, aes(Lfgm,Wfgm))
graph.a
```

## Adding Geoms: `geom_point()`
```{r, fig.align='center', fig.height=2.75}
graph.a + geom_point()
```

## Adding Geoms: `geom_smooth()`
```{r, fig.align='center', fig.height=2.75}
graph.a + geom_point() + geom_smooth(method = 'loess')
```

## Adding Geoms: `geom_rug()`
```{r, fig.align='center', fig.height=2.75}
graph.a + geom_point() + geom_smooth(method = 'loess') + geom_rug()
```

## Adding Geoms: `geom_density2d()`
```{r, fig.align='center', fig.height=2.75}
graph.a + geom_point() + geom_smooth(method = 'loess') +
  geom_rug() + geom_density2d()
```

## Adding Geoms: `geom_jitter()`
```{r, fig.align='center', fig.height=2.75}
graph.a + geom_rug() + geom_density2d() + geom_jitter()
```

## Adding Geoms: `labs()`
```{r, fig.align='center', fig.height=2.65}
graph.a  + geom_rug() + geom_density2d() +
 geom_jitter() + 
  labs(x='Losing Team Field Goals Made', 
       y = 'Winning Team Field Goals Made')
```

## Scales: `xlim()` and `ylim()`
```{r, mysize=TRUE, fig.align='center',warning=FALSE, fig.height=2.6}
graph.a + geom_rug() + geom_density2d() +
 geom_jitter() + 
  labs(x='Losing Team Field Goals Made', 
       y = 'Winning Team Field Goals Made') +
  xlim(c(0,max(hoops_2011$Wfgm))) + ylim(c(0,max(hoops_2011$Wfgm)))
```

## Themes
There are a wide range of themes available in ggplot: [theme overview](https://ggplot2.tidyverse.org/reference/ggtheme.html)
```{r,fig.align='center', fig.height=2.75, echo = F}
graph.a + geom_point() + theme_light() + labs(x='Losing Team Field Goals Made', y = 'Winning Team Field Goals Made')
```

## More about aes
```{r, fig.align='center', fig.height=2.75}
graph.a + geom_jitter(col = 'firebrick4')
```

## More about aes
```{r, fig.align='center', fig.height=2.75}
graph.a + geom_jitter(aes(col = as.factor(Season)))
```


## More about aes
```{r, fig.align='center', fig.height=2.75}
graph.a + geom_jitter(aes(col = as.factor(Season)), size=3,alpha=.4)
```

## More about aes
```{r, echo = F, mysize=TRUE,fig.align='center'}
graph.a + geom_jitter(aes(shape = as.factor(Season),col=Wscore), size=3,alpha=.4)
```

## More about aes: Comment
```{r, eval = F, mysize=TRUE,fig.align='center'}
graph.a + 
  geom_jitter(aes(shape = as.factor(Season),col=Wscore),
              size=3,alpha=.4)
```

## Faceting
```{r, echo = F}
graph.a + geom_point() + facet_wrap(~Season)
```

## Faceting: Comment
```{r, eval = F}
graph.a + geom_point() + facet_wrap(~Season)
```

## Faceting
```{r, mysize=TRUE, eval=T, size='\\tiny',fig.align='center', fig.height=2.75}
graph.a + facet_wrap(~Season) + geom_jitter(alpha=.5, aes(color=Wfgm3))
```

## Exercise: ggplot2

Now use `ggplot2` to create an interesting graph using the Seattle Housing data set.

## Solution: ggplot2
```{r,fig.align='center',echo=F}
seattle_in$zipcode <- as.factor(seattle_in$zipcode)
graph.a <- ggplot(data = seattle_in, aes(sqft_living,price))
graph.a + geom_jitter(aes(col = zipcode))+ theme(plot.title = element_text(size=8), text = element_text(size=6)) + geom_smooth(method='loess')+ggtitle('Seattle Housing Sales: Price vs. Square Footage Living Space') + ylab('Sales Price (million dollars)') + xlab('Living Space (square foot)')+scale_y_continuous(breaks=c(seq(0,7000000,by=1000000)), labels=as.character(0:7)) +  annotate('text',3500,6000000, label = 'Housing price depends on zipcode', size=2) +annotate("rect", xmin = 0, xmax = 7250, ymin = 5500000, ymax = 6500000, alpha = .6) + geom_segment(aes(x=3500, xend=3500, y=5500000, yend=3000000),
                           arrow = arrow(length = unit(0.5, "cm")))
```

## Solution: ggplot2
```{r, eval = F}
seattle_in$zipcode <- as.factor(seattle_in$zipcode)
graph.a <- ggplot(data = seattle_in, aes(sqft_living,price))
graph.a + geom_jitter(aes(col = zipcode))+ 
  theme(plot.title = element_text(size=8), 
        text = element_text(size=6)) + geom_smooth(method='loess')+
  ggtitle('Seattle Housing Sales: Price vs. Square Footage Living Space') + 
  ylab('Sales Price (million dollars)') + 
  xlab('Living Space (square foot)')+
  scale_y_continuous(breaks=c(seq(0,7000000,by=1000000)),
                     labels=as.character(0:7)) +  
  annotate('text',3500,6000000, 
           label = 'Housing price depends on zipcode', size=2) +
  annotate("rect", xmin = 0, xmax = 7250, ymin = 5500000, ymax = 6500000, alpha = .6) + 
  geom_segment(aes(x=3500, xend=3500, y=5500000, yend=3000000),
                           arrow = arrow(length = unit(0.5, "cm")))
```
