---
title: "Polished Graphics"
format: html
---

```{r setup, include=FALSE}
library(tidyverse)
```

#### Graphics that Stand Alone

__From now on, all graphics must "stand alone".__ This means informative titles, axes, labels, captions, annotations - when appropriate,  etc...

#### Bar Charts with Count Data / Statistical Transformation

```{r read_seattle, message = F}
seattle <- read_csv('http://math.montana.edu/ahoegh/teaching/stat408/datasets/SeattleHousing.csv') |>
  mutate(zipcode = factor(zipcode))
```

In @fig-bar1 we see a bar chart made without using tally to count up the number of houses sold in each zipcode

```{r bar_plot1, fig.cap = "Standard Bar Chart without tally()"}
#| label: fig-bar1
seattle |>
  ggplot(aes(x = zipcode, fill = zipcode)) +
  geom_bar() + 
  labs(title = 'Figure 1. First Bar Plot: # of houses sold in King County, WA zipcodes') +
  ylab("number of houses") +
  theme_minimal() +
  theme(legend.position = 'none') 
  
```

#### Activity #1

Recreate this figure using `seattle_tallied` and use `library(patchwork)` to combine the images into a single figure.

```{r}
seattle_tallied <- seattle |>
  group_by(zipcode) |>
  tally()
```

```{r figure2}
#| label: fig-bar2
#| fig-cap: "One plot, two ways"
#| fig-height: 8
library(patchwork)
f1 <- seattle |>
  ggplot(aes(x = zipcode, fill = zipcode)) +
  geom_bar() + 
  labs(title = 'Figure 2a. First Bar Plot: # of houses sold in King County, WA zipcodes') +
  ylab("number of houses") +
  theme_minimal() +
  theme(legend.position = 'none') 

f2 <- seattle_tallied |>
  ggplot(aes(x = zipcode, y = n, fill = zipcode)) +
  geom_bar(stat = "identity") + 
  labs(title = 'Figure 2b. First Bar Plot: # of houses sold in King County, WA zipcodes') +
  ylab("number of houses") +
  theme_minimal() +
  theme(legend.position = 'none') 

f1 / f2
```

#### Activity #2

Now expand on @fig-bar1 and also include waterfront in this figure. One option using faceting is shown in @fig-bar3.


```{r}
#| label: fig-bar3
#| fig-cap: "Ugly alert"
seattle |>
  ggplot(aes(x = zipcode, fill = zipcode)) +
  geom_bar() + 
  labs(title = 'Figure 3. Faceted Bar Plot: # of houses sold in King County, WA zipcodes') +
  ylab("number of houses") +
  theme_minimal() +
  theme(legend.position = 'none') +
  facet_wrap(waterfront ~ ., nrow = 2)
```

A better approach in this case would be a stacked bar plot as in @fig-bar4. You could also do a side-by-side bar chart using `geom_bar(position = 'dodge')`


```{r}
#| label: fig-bar4
#| fig-cap: "Better than @fig-bar3"
seattle |>
  mutate(waterfront = factor(waterfront)) |>
  ggplot(aes(x = zipcode, color = waterfront)) +
  geom_bar() + 
  labs(title = 'Figure 4. Stacked Bar Plot: # of houses sold in King County, WA zipcodes') +
  ylab("number of houses") +
  theme_minimal() +
  theme(legend.position = 'bottom') 
```

---

#### On bar charts

- Bar charts are good for summarizing counts (better than tables)

- Bar charts are bad at showing variability in the data...

```{r}
#| label: fig-compare
#| fig-cap: "Boxplot shows much more information, with less ink"
#| fig-height: 8

bar1 <- seattle |>
  group_by(zipcode) |>
  summarise(price = mean(price)) |>
  ggplot(aes(x = zipcode, y = price, fill = zipcode)) +
  geom_bar(stat = 'identity') + 
  labs(title = 'Figure 5. Bad Bar Chart \n # of houses sold in King County, WA zipcodes') +
  ylab("Average Sales Price") +
  theme_minimal() +
  theme(legend.position = 'none') 

box1 <- seattle |>
  ggplot(aes(x = zipcode, y = price, fill = zipcode)) +
  geom_boxplot() + 
  labs(title = 'Figure 5. Better Box Plot \n # of houses sold in King County, WA zipcodes') +
  ylab("Average Sales Price") +
  theme_minimal() +
  theme(legend.position = 'none') 


bar1 / box1
```


#### Activity #3

Update @fig-compare so that the y label is easy to interpret (in dollars).

```{r, message = F}
#| label: fig-compare2
#| fig-cap: "Boxplot shows much more information, with less ink"
#| fig-height: 8
library(scales)
bar1 <- seattle |>
  group_by(zipcode) |>
  summarise(price = mean(price)) |>
  ggplot(aes(x = zipcode, y = price, fill = zipcode)) +
  geom_bar(stat = 'identity') + 
  labs(title = 'Figure 6. Bad Bar Chart \n # of houses sold in King County, WA zipcodes') +
  ylab("Average Sales Price") +
  theme_minimal() +
  theme(legend.position = 'none') +
scale_y_continuous(labels = dollar)

box1 <- seattle |>
  ggplot(aes(x = zipcode, y = price, fill = zipcode)) +
  geom_boxplot() + 
  labs(title = 'Figure 6. Better Box Plot \n # of houses sold in King County, WA zipcodes') +
  ylab("Average Sales Price") +
  theme_minimal() +
  theme(legend.position = 'none') +
  scale_y_continuous(labels = dollar)


bar1 / box1
```

#### Activity #4

Make the y-axes on both plots the same.

Update @fig-compare so that the y label is easy to interpret (in dollars).

```{r, message = F}
#| label: fig-compare3
#| fig-cap: "Adjusting axis limits and adding annotation"
#| fig-height: 8
library(scales)

fancy_pants <- seattle |>
  filter(price > 4000000)

bar1 <- seattle |>
  group_by(zipcode) |>
  summarise(price = mean(price)) |>
  ggplot(aes(x = zipcode, y = price, fill = zipcode)) +
  geom_bar(stat = 'identity') + 
  labs(title = 'Figure 7. Bad Bar Chart \n # of houses sold in King County, WA zipcodes') +
  ylab("Average Sales Price") +
  theme_minimal() +
  theme(legend.position = 'none') + 
scale_y_continuous(labels = dollar, limits = c(NA, 8000000)) +
  annotate('text', x = 3, y = 6400000, label = 'fancy pants houses up here') +
  geom_curve(
  aes(x = 3, y = 6000000, xend = 5, yend = 7000000),
  arrow = arrow(
    length = unit(0.03, "npc"), type="closed" ), ) +
  labs(caption = 'bar chart only shows 1 piece of information - the mean') +
  geom_text(data = fancy_pants, label = 'Expensive!') 


box1 <- seattle |>
  ggplot(aes(x = zipcode, y = price, fill = zipcode)) +
  geom_boxplot() + 
  labs(title = 'Figure 7. Better Box Plot \n # of houses sold in King County, WA zipcodes') +
  ylab("Average Sales Price") +
  theme_minimal() +
  theme(legend.position = 'none') +
  scale_y_continuous(labels = dollar)


bar1 / box1
```

## Accessibility (color blindness)

Code from # Code from [https://github.com/clauswilke/colorblindr](https://github.com/clauswilke/colorblindr)


```{r}
fig <- ggplot(iris, aes(Sepal.Length, fill = Species)) + geom_density(alpha = 0.7)
fig
```

```{r}
library(remotes)
#remotes::install_github("clauswilke/colorblindr")
library(colorblindr)
cvd_grid(fig)

```
#### Activity #5

Improve this plot of price vs sqft_living (with bedrooms). Add titles, labels, and use a colorblind friendly way to highlight the number of bedrooms in a house. Also explore different themes.

```{r}
seattle |>
  ggplot(aes(y = price, x= sqft_living, color = bedrooms)) +
  geom_point()
```

```{r}
#| fig-height: 8
library(viridis)
seattle |>
  ggplot(aes(y = price, x= sqft_living, color = bedrooms)) +
  geom_point() +
  scale_color_viridis() +
  facet_wrap(.~zipcode) +
  theme(legend.position = 'bottom') +
  theme_dark() +
  xlab(expression(paste("Living space (",ft^2,')',sep=""))) +
  labs('House price by living space in King County, WA')
```

```{r}
#| fig-height: 30
library(viridis)
seattle |>
  mutate(bedrooms = factor(bedrooms)) |>
  ggplot(aes(y = price, x= sqft_living, color = bedrooms)) +
  geom_point() +
  scale_color_viridis(discrete = T) +
  facet_wrap(.~zipcode, ncol = 1) +
  theme_minimal() +
  xlab(expression(paste("Living space (",ft^2,')',sep=""))) +
  labs('House price by living space in King County, WA') +
  theme(legend.position = 'bottom') +
  guides(colour = guide_legend(nrow = 1))

```

### Additional Color Palletes

```{r }
#| fig-height: 10
a <- seattle |>
  ggplot(aes(x = zipcode, fill = zipcode)) +
  geom_bar() + 
  labs(title = 'Figure X. More Color Palletes') +
  ylab("number of houses") +
  theme_minimal() +
  theme(legend.position = 'none',
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_fill_viridis(discrete = T) 


b <- seattle |>
  ggplot(aes(x = zipcode, fill = zipcode)) +
  geom_bar() + 
  labs(title = 'Figure X. More Color Palletes') +
  ylab("number of houses") +
  theme_minimal() +
  theme(legend.position = 'none',
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_fill_viridis(discrete = TRUE, option = 'B')
  
c <- seattle |>
  ggplot(aes(x = zipcode, fill = zipcode)) +
  geom_bar() + 
  labs(title = 'Figure X. More Color Palletes') +
  ylab("number of houses") +
  theme_minimal() +
  theme(legend.position = 'none',
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_fill_viridis(discrete = TRUE, option = 'C')

d <- seattle |>
  ggplot(aes(x = zipcode, fill = zipcode)) +
  geom_bar() + 
  labs(title = 'Figure X. More Color Palletes') +
  ylab("number of houses") +
  theme_minimal() +
  theme(legend.position = 'none',
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_fill_viridis(discrete = TRUE, option = 'D')

e <- seattle |>
  ggplot(aes(x = zipcode, fill = zipcode)) +
  geom_bar() + 
  labs(title = 'Figure X. More Color Palletes') +
  ylab("number of houses") +
  theme_minimal() +
  theme(legend.position = 'none',
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_fill_viridis(discrete = TRUE, option = 'E')


(a + b) / (c + d) / e
```
